# Installing software

## R Packages 

These are the packages used throughout this project.


```{r, echo=TRUE, eval = FALSE }
library(sf)           # To work with geospatial data
library(raster)       # To work with Raster data
library(glue)         # To work with text data
library(gtfstools)    # To work with GTFS data
library(tidyverse)    # To load the different data wrangling packages
library(lubridate)    # To work with dates
library(colorspace)   # For extra color options
library(terra)        # To work with Raster data
library(opentripplanner) # to access the OpenTripPlanner API
library(gtfs2gps)     # To work with GTFS data
library(elevatr)      # To download elevation data
library(stringi)      # To work with text data
library(openrouteservice) # to access the openrouteservice API
library(mapview)     # To create interactive maps
library(ggspatial)   # To create static maps
```

These are the packages and package versions that were used in this project.

```{r, echo=TRUE, eval=FALSE}
sessionInfo() 
```

```{r, echo=TRUE, eval=FALSE}
R version 4.1.0 (2021-05-18)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 10 x64 (build 22621)

Matrix products: default

locale:
[1] LC_COLLATE=English_United States.1252  LC_CTYPE=English_United States.1252    LC_MONETARY=English_United States.1252 LC_NUMERIC=C                          
[5] LC_TIME=English_United States.1252    

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] ggspatial_1.1.6        mapview_2.11.0         openrouteservice_0.4.1 stringi_1.6.2          elevatr_0.4.2          gtfs2gps_2.0-2         opentripplanner_0.4.0 
 [8] terra_1.5-21           colorspace_2.0-3       lubridate_1.7.10       forcats_0.5.1          stringr_1.4.0          dplyr_1.0.7            purrr_0.3.4           
[15] readr_2.0.0            tidyr_1.1.3            tibble_3.1.7           ggplot2_3.3.6          tidyverse_1.3.1        gtfstools_1.0.0        glue_1.6.2            
[22] raster_3.5-15          sp_1.4-7               sf_1.0-1              

loaded via a namespace (and not attached):
 [1] leafem_0.1.6        ellipsis_0.3.2      class_7.3-19        leaflet_2.1.1       satellite_1.0.4     base64enc_0.1-3     fs_1.5.0            xaringanExtra_0.5.4
 [9] rstudioapi_0.13     httpcode_0.3.0      proxy_0.4-26        fansi_1.0.3         xml2_1.3.3          codetools_0.2-18    knitr_1.39          geojsonlint_0.4.0  
[17] jsonlite_1.8.0      broom_0.7.8         dbplyr_2.1.1        png_0.1-7           compiler_4.1.0      httr_1.4.3          keyring_1.3.0       backports_1.2.1    
[25] assertthat_0.2.1    fastmap_1.1.0       lazyeval_0.2.2      cli_3.3.0           htmltools_0.5.2     tools_4.1.0         gtable_0.3.0        ggthemes_4.2.4     
[33] V8_4.1.0            Rcpp_1.0.8.3        cellranger_1.1.0    vctrs_0.4.1         crul_1.2.0          svglite_2.0.0       progressr_0.10.0    crosstalk_1.2.0    
[41] lwgeom_0.2-6        xfun_0.30           rvest_1.0.0         lifecycle_1.0.1     scales_1.2.0        hms_1.1.0           parallel_4.1.0      yaml_2.3.5         
[49] curl_4.3.2          pbapply_1.5-0       pander_0.6.4        jsonvalidate_1.3.2  e1071_1.7-7         rlang_1.0.2         pkgconfig_2.0.3     systemfonts_1.0.2  
[57] evaluate_0.15       lattice_0.20-44     htmlwidgets_1.5.4   tidyselect_1.1.1    magrittr_2.0.3      bookdown_0.27       geojsonsf_2.0.2     R6_2.5.1           
[65] generics_0.1.0      DBI_1.1.1           pillar_1.7.0        haven_2.4.1         withr_2.5.0         units_0.7-2         jtools_2.2.0        modelr_0.1.8       
[73] crayon_1.5.1        KernSmooth_2.23-20  utf8_1.2.2          plotly_4.9.4.1      tzdb_0.1.2          rmarkdown_2.14      grid_4.1.0          readxl_1.3.1       
[81] data.table_1.14.0   reprex_2.0.0        digest_0.6.29       classInt_0.4-3      webshot_0.5.3       stats4_4.1.0        munsell_0.5.0       viridisLite_0.4.0  
[89] kableExtra_1.3.4   
```



## OSMConvert

To speed up the process of OpenTripPlanner and the routing algorithm, it is helpful if the OSM map is as small as possible, while still containing the whole FUA. A tool that allows you to ‘crop’ the OSM map to desired size is OSMConvert^[https://wiki.openstreetmap.org/wiki/Osmconvert]. The [OpenTripPlanner website](http://docs.opentripplanner.org/en/latest/Preparing-OSM/) and the [OpenStreetMap Wiki](https://wiki.openstreetmap.org/wiki/Osmconvert) have good documentation on how to use the tool. As it is an executable, you can directly run it from R  using the `system()` function. 


## OpenTripPlanner


The main software that is used for the routing is called OpenTripPlanner^[https://www.opentripplanner.org/]. *"OpenTripPlanner (OTP) is a family of open source software projects that provide passenger information and transportation network analysis services. The core server-side Java component finds itineraries combining transit, pedestrian, bicycle, and car segments through networks built from widely available, open standard OpenStreetMap and GTFS data. This service can be accessed directly via its web API or using a range of Javascript client libraries, including modern reactive modular components targeting mobile platforms."*

To install OpenTripPlanner, you can use the `opentripplanner` R package.^[https://cran.r-project.org/web/packages/opentripplanner/vignettes/opentripplanner.html]

```{r, eval=FALSE}
install.packages("opentripplanner") # Install Package
library(opentripplanner)            # Load Package
```

For more information on how to use this package, please refer to [vignette and the starters guide](https://docs.ropensci.org/opentripplanner/articles/opentripplanner.html#getting-data-for-otp)


##  OpenRouteService API

The  OpenRouteService API^[https://openrouteservice.org/] does not differ too much in its use from OpenTripPlanner. Both run locally and are based on OpenStreetMaps. OpenRouteService leverages the use of the Dijskta Algorithm and because it is implemented in C++ and runs via Docker, which makes it very fast to use. The OpenRouteService API also includes an option to compute a walking isodistance polygon, something that is not straight-forward in OpenTripPlanner. To use the  OpenRouteService API, you need to install Docker desktop.^[https://www.docker.com/products/docker-desktop/] You can then pull the openroute service container^[https://hub.docker.com/r/giscience/openrouteservice] from Docker, using this command in the Docker command prompt.

```{bash eval=FALSE}
docker pull giscience/openrouteservice
```

For a more detailed description of how to set up the Docker image, see the  [helpfiles of OpenRouteService](https://giscience.github.io/openrouteservice/installation/Installation-and-Usage.html) or this very helpful blog by Dipayan Banerjee^[(Setting up openrouteservice via Docker for Windows 10 Home)[https://sites.google.com/view/dipayan-banerjee/local-openrouteservice-setup-windows-10-home]]. By default, the openrouteservice docker container only builds the profile necessary to compute driving routes. If you want to add other profiles (for example, the profile needed for computing the optimal walking route), you will have to make some adjustments in the `ors-config.json` file. 




